Follow-up Samples

### 2017-05-03 submission/manuscript.Rmd
put together Mack's changes

Building upon this evidence, several human studies have identified unique signatures of colonic lesions [@wang_structural_2012; @chen_decreased_2013; @chen_human_2012; @shen_molecular_2010; @kostic_genomic_2012; @feng_gut_2015]. One line of research has identified community-level differences between those bacteria that are found on and adjacent to colonic lesions and have supported a role for *Bacteroides fragilis*, *Escherichia coli*, and *Fusobacterium nucleatum* in tumorigenesis [@dejea_microbiota_2014; @mima_fusobacterium_2015; @arthur_intestinal_2012]. Others have proposed feces-based biomarkers that could be used to diagnose the presence of colonic adenomas and carcinomas [@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014]. These studies have associated *Fusobacterium nucleatum* and other oral pathogens with colonic lesions (adenoma, advanced adenoma, and carcinoma). They have also noted that the loss of bacteria generally thought to produce short chain fatty acids, which can suppress inflammation, is associated with colonic lesions. This suggests that gut bacteria have a role in tumorigenesis with potential as useful biomarkers for aiding in the early detection of disease.

Despite advances in understanding the role between the gut microbiota and colonic tumorigenesis, we still do not understand how treatments including resection, chemotherapy, and/or radiation affect the composition of the gut microbiota. If the community drives tumorigenesis then one would hypothesize that treatment to remove a lesion would affect the microbiota and risk of recurrence. To test this hypothesis, we addressed two related questions: Does treatment affect the colonic microbiota in a predictable manner? If so, does the treatment alter the community to more closely resemble that of individuals with normal colons?

@@

***The effects of treatment are not consistent across treatment groups.***
We used two approaches to identify those bacterial populations that change between the two samples for each diagnosis group. First, we sought to identify individual OTUs that could account for the change in overall community structure. However, using a paired Wilcoxon test we were unable to identify any OTUs that were significantly different in the pre and post-treatment groups (P-value `r ifelse(all_OTU_pair_wilcox > 0.05, invisible(">"), invisible("<"))` 0.05). It is likely that high inter-individual variation and the irregular distribution of OTUs across individuals limited the statistical power of the test. To overcome these problems we developed Random Forest models to identify collections of OTUs that would allow us to differentiate between pre and post-treatment samples from each of the diagnosis groups. Because of the relatively small number of subjects in each group, we restricted our models to only incorporate 10 OTUs to limit the likelihood that the models would overfit the data. Despite this restriction, the models performed well (adenoma AUC range = `r format(min(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`). Interestingly, the 10 OTUs that were used for each model had little overlap with each other [Figure 2]. These results support the earlier community-wide analysis where we observed that the treatment had an impact on the overall community structure; however, the effect of treatment was not consistent across patients and diagnosis groups.

@@

This study builds upon previous work that demonstrated a role for the microbiota in tumorigenesis and that microbiome-derived biomarkers can detect colonic lesions [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_microbial_2014; @kostic_fusobacterium_2013; @wu_human_2009; @yu_metagenomic_2017; @baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014; @warren_co-occurrence_2013]. Our study focused on comparing the microbiota of patients diagnosed with adenoma, advanced adenoma, and carcinoma before and after treatment. For all three groups of patients, we observed changes in their microbiota. After treatment, the microbiota of patients with carcinoma changed significantly more than the other groups. This change resulted in communities that more closely resembled those of patients with a normal colon.  This may suggest that treatment for carcinoma is not only successful for removing the carcinoma but also at reducing the associated bacterial communities.  Understanding the effect of treatment on the microbiota of those diagnosed with carcinomas may have important implications for reducing disease recurrence.  It is intriguing that it may be possible to use microbiome-based biomarkers to not only predict the presence of lesions but to assess the risk of recurrence.

@@

There was a subset of patients (`r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` of the `r length(rownames(filter(good_metaf, dx == "cancer")))` with carcinomas) who demonstrated an elevated probability of carcinoma after treatment. This may reflect an elevated risk of recurrence. The `r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`% prevalence of increased carcinoma probability from our study is within the expected rate of recurrence (20-30% [@hellinger_reoperation_2006; @ryuk_predictive_2014]). We hypothesized that these individuals may have had more severe tumors; however, the tumor severity of these `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` individuals (`r table(good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes) %>% select(stage))[1]` with Stage II and `r table(good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes) %>% select(stage))[2]` with Stage III) was similar to the distribution observed among the other `r length(rownames(filter(good_metaf, dx == "cancer"))) - table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` patients. We also hypothesized that we may have sampled these patients later than the rest and their communities may have reverted to a carcinoma-associated state; however, there was not a statistically significant difference in the length of time between sample collection among those whose probabilities increased or decreased (Wilcoxon Test; P-value = `r format(wilcox.test((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"], (good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"])$p.value, digits = 2)`). Finally, it is possible that these patients may not have responded to treatment as well as the other `r length(rownames(filter(good_metaf, dx == "cancer"))) - table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` patients diagnosed with carcinoma and so the microbiota may not have been impacted the same way. Again, further studies looking at the role of the microbiota in recurrence are needed to understand the dynamics following treatment.

Our final hypothesis was that the specific type of treatment altered the structure of the microbiome. The treatment to remove adenomas and advanced adenomas was either polyp removal or surgical resection whereas it was surgical resection alone or in combination with chemotherapy or with chemotherapy and radiation for individuals with carcinoma. Because chemotherapy and radiation target rapidly growing cells, these treatments would be more likely to cause a turnover of the colonic epithelium driving a more significant change in the structure of the microbiota. Although, we were able to test for an effect across these specific types of treatment, the number of patients in each treatment group was relatively small.

@@

```{r table1, echo=FALSE, message=FALSE, warning=FALSE}
test_cohort <- good_metaf %>% select(Dx_Bin, dx, Age, Gender, BMI, White)

t2 <- as.data.frame(matrix(nrow = 5, ncol = 3, 
                           dimnames = 
                             list(rown = c("Dx_Bin", "Age", "Gender_f", "BMI", "White"), 
                                  coln = c("Adenoma", "SRN", "Carcinoma"))))

@@

  t2["Dx_Bin", i] <- length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i])))
  t2["Age", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(Age))
  t2["Gender_f", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% count(Gender) %>% mutate(n = n/sum(n) * 100) %>%  filter(Gender == "f") %>% select(n)
  t2["BMI", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(BMI, na.rm = TRUE))
  t2["White", i] <- length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & White == 1)))/length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]))) * 100 
}

@@

  age_sd <- c(age_sd, (filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(Age)))[, "std"])
  BMI_sd <- c(BMI_sd, (filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(BMI, na.rm = TRUE)))[, "std"])
}

@@

ttemptable <- as.data.frame(t(t2)) %>% mutate(Age = paste(format(Age, digits = 2, nsmall = 2), 
                                                    label = "\u00B1", 
                                                    format(age_sd, digits = 2, nmsall = 2)), 
                                        Gender_f = format(Gender_f, digits = 2, nsmall = 2), 
                                        BMI = paste(format(BMI, digits = 2, nsmall = 2), 
                                                    label = "\u00B1", 
                                                    format(BMI_sd, digits = 2, nmsall = 2)), 
                                        White = format(White, digits =2, nsmall = 2))

ft1 <- cbind(Group = c("Adenoma", "Advanced Adenoma", "Carcinoma"), ttemptable$Dx_Bin, ttemptable$Age, ttemptable$Gender_f, 
             ttemptable$BMI, ttemptable$White)

>>>>

Building upon this evidence, several human studies have identified unique signatures of colonic lesions [@wang_structural_2012; @chen_decreased_2013; @chen_human_2012; @shen_molecular_2010; @kostic_genomic_2012; @feng_gut_2015]. One line of research has identified community-level differences between those bacteria that are found on and adjacent to colonic lesions and have supported a role for *Bacteroides fragilis*, *Escherichia coli*, and *Fusobacterium nucleatum* in tumorigenesis [@dejea_microbiota_2014; @mima_fusobacterium_2015; @arthur_intestinal_2012]. Others have proposed feces-based biomarkers that could be used to diagnose the presence of colonic adenomas and carcinomas [@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014]. These studies have associated *Fusobacterium nucleatum* and other oral pathogens with colonic lesions (adenoma, advanced adenoma, and carcinoma). They have also noted that the loss of bacteria generally thought to produce short chain fatty acids, which can suppress inflammation, is associated with colonic lesions. This suggests that gut bacteria have a role in tumorigenesis with potential as useful biomarkers for aiding in the early detection of disease [@dejea_microbiota_2014; @mima_fusobacterium_2015; @arthur_intestinal_2012;@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014].

Despite advances in understanding the role between the gut microbiota and colonic tumorigenesis, we still do not understand how treatments including resection, chemotherapy, and/or radiation affect the composition of the gut microbiota. If the microbial community drives tumorigenesis then one would hypothesize that treatment to remove a lesion would affect the microbiota and risk of recurrence. To test this hypothesis, we addressed two related questions: Does treatment affect the colonic microbiota in a predictable manner? If so, does the treatment alter the community to more closely resemble that of individuals with normal colons?

@@

***The effects of treatment are not consistent across treatment groups.***
We used two approaches to identify those bacterial populations that change between the two samples for each diagnosis group. First, we sought to identify individual OTUs that could account for the change in overall community structure. However, using a paired Wilcoxon test we were unable to identify any OTUs that were significantly different in the pre and post-treatment groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected) > 0.05, invisible(">"), invisible("<"))` 0.05). It is likely that high inter-individual variation and the irregular distribution of OTUs across individuals limited the statistical power of the test. To overcome these problems we developed Random Forest models to identify collections of OTUs that would allow us to differentiate between pre and post-treatment samples from each of the diagnosis groups. Because of the relatively small number of subjects in each group, we restricted our models to only incorporate 10 OTUs to limit the likelihood that the models would overfit the data. Despite this restriction, the models performed well (adenoma AUC range = `r format(min(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`). Interestingly, the 10 OTUs that were used for each model had little overlap with each other [Figure 2]. These results support the earlier community-wide analysis where we observed that the treatment had an impact on the overall community structure; however, the effect of treatment was not consistent across patients and diagnosis groups.

@@

Our study focused on comparing the microbiota of patients diagnosed with adenoma, advanced adenoma, and carcinoma before and after treatment. For all three groups of patients, we observed changes in their microbiota. After treatment, the microbiota of patients with carcinoma changed significantly more than the other groups. This change resulted in communities that more closely resembled those of patients with a normal colon.  This may suggest that treatment for carcinoma is not only successful for removing the carcinoma but also at reducing the associated bacterial communities.  Understanding the effect of treatment on the microbiota of those diagnosed with carcinomas may have important implications for reducing disease recurrence.  It is intriguing that it may be possible to use microbiome-based biomarkers to not only predict the presence of lesions but to assess the risk of recurrence.

@@

There was a subset of patients (`r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` of the `r length(rownames(filter(good_metaf, dx == "cancer")))` with carcinomas) who demonstrated an elevated probability of carcinoma after treatment. This may reflect an elevated risk of recurrence. The `r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`% prevalence of increased carcinoma probability from our study is within the expected rate of recurrence (20-30% [@hellinger_reoperation_2006; @ryuk_predictive_2014]). We hypothesized that these individuals may have had more severe tumors; however, the tumor severity of these `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` individuals (`r table(good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes) %>% select(stage))[1]` with Stage II and `r table(good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes) %>% select(stage))[2]` with Stage III) was similar to the distribution observed among the other `r length(rownames(filter(good_metaf, dx == "cancer"))) - table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` patients. We also hypothesized that we may have sampled these patients later than the rest and their communities may have reverted to a carcinoma-associated state; however, there was not a statistically significant difference in the length of time between sample collection among those whose probabilities increased (`r format(mean((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"]), digits = 0)` $\pm$ `r format(sd((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"]), digits = 0)` days) or decreased (`r format(mean((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"]), digits = 0)` $\pm$ `r format(sd((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"]), digits = 0)` days) (Wilcoxon Test; P-value = `r format(wilcox.test((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"], (good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"])$p.value, digits = 2)`). Finally, it is possible that these patients may not have responded to treatment as well as the other `r length(rownames(filter(good_metaf, dx == "cancer"))) - table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` patients diagnosed with carcinoma and so the microbiota may not have been impacted the same way. Again, further studies looking at the role of the microbiota in recurrence are needed to understand the dynamics following treatment.

Our final hypothesis was that the specific type of treatment altered the structure of the microbiome. The treatment to remove adenomas and advanced adenomas was either polyp removal or surgical resection whereas it was surgical resection alone or in combination with chemotherapy or with chemotherapy and radiation for individuals with carcinoma. Because chemotherapy and radiation target rapidly growing cells, these treatments would be more likely to cause a turnover of the colonic epithelium driving a more significant change in the structure of the microbiota. Although, we were able to test for an effect across these specific types of treatment, the number of patients in each treatment group was relatively small.  Finally, those undergoing surgery would have received antibiotics and this may be a potential confounder.  However, our pre-treatment stool samples were obtained before the surgery and the post-treatment samples were obtained long after any effects due to antibiotic administration on the microbiome would be expected to occur (`r format(mean(filter(good_metaf, dx == "cancer")[, "time"]), digits = 0)` $\pm$ `r format(sd(filter(good_metaf, dx == "cancer")[, "time"]), digits = 0)` days).

@@

```{r table1, echo=FALSE, message=FALSE, warning=FALSE}
test_cohort <- good_metaf %>% select(Dx_Bin, dx, Age, Gender, BMI, White, time)

t2 <- as.data.frame(matrix(nrow = 6, ncol = 3, 
                           dimnames = 
                             list(rown = c("Dx_Bin", "Age", "Gender_f", "BMI", "White", "post_time"), 
                                  coln = c("Adenoma", "SRN", "Carcinoma"))))

@@

  t2["Dx_Bin", i] <- length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i])))
  t2["Age", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(Age))
  t2["Gender_f", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% count(Gender) %>% mutate(n = n/sum(n) * 100) %>%  filter(Gender == "f") %>% select(n)
  t2["BMI", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(BMI, na.rm = TRUE))
  t2["White", i] <- length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & White == 1)))/length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]))) * 100

  t2["post_time", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(time, na.rm = TRUE))

}

# Generate SD for applicable metrics
age_sd <- c()
BMI_sd <- c()
post_time_sd <- c()

@@

  age_sd <- c(age_sd, (filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(Age)))[, "std"])
  BMI_sd <- c(BMI_sd, (filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(BMI, na.rm = TRUE)))[, "std"])
  post_time_sd <- c(post_time_sd, (filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(time, na.rm = TRUE)))[, "std"])

}

@@

ttemptable <- as.data.frame(t(t2)) %>% mutate(Age = paste(format(Age, digits = 2, nsmall = 2), 
                                                    label = "\u00B1", 
                                                    format(age_sd, digits = 2, nmsall = 2)), 
                                        Gender_f = format(Gender_f, digits = 2, nsmall = 2), 
                                        BMI = paste(format(BMI, digits = 2, nsmall = 2), 
                                                    label = "\u00B1", 
                                                    format(BMI_sd, digits = 2, nmsall = 2)), 
                                        White = format(White, digits =2, nsmall = 2), 
                                        post_time = paste(format(post_time, digits = 2, nsmall = 2), 
                                                    label = "\u00B1", 
                                                    format(post_time_sd, digits = 2, nmsall = 2)))

ft1 <- cbind(Group = c("Adenoma", "Advanced Adenoma", "Carcinoma"), ttemptable$Dx_Bin, ttemptable$Age, ttemptable$Gender_f, 
             ttemptable$BMI, ttemptable$White, ttemptable$post_time)

age_label <- paste("Age (Mean ", label = "\u00B1", " SD)", sep = "")
bmi_label <- paste("BMI (Mean ", label = "\u00B1", " SD)", sep = "")
post_time_label <- paste("Post-Treatment Sample (Mean ", label = "\u00B1", " SD)", sep = "")

colnames(ft1) <- c("Group", "n", age_label, "Sex (%F)", bmi_label, "Caucasian (%)", post_time_label)
